Optimising Heterogeneous Ambulance Fleet Allocations in Jakarta

### 2024-08-12 response/response.tex

\begin{itemize}

\item Reviewer 2 wrote:
\begin{quote}
``The introduction makes a case for using survival vs reponse time as a measure for ambulance allocation (although this is not the only measure in the results). What would this do to fairness in the allocations? Would far away patients with serious conditions (ie. Needing a fast response time) just be discarded as it is expensive to cater for them (in terms of overall survival of the population)?''
\end{quote}
A paragraph has now been added to Section 4.1 comparing allocations that maximise survival, maximise coverage, and minimise response times.

\item Reviewer 2 wrote:
\begin{quote}
``Is the survival function the same for $K_a$ patients regardless of which vehicle visits them? What is the rationale for this? It seems that primary vehicles would be better prepared to deal with complications, etc.''
\end{quote}
This is an interesting suggestion. Currently we model the survival function to be independent of the vehicle type, that is $s_k$ is the same regardless if the vehicle that arrived is a primary or secondary vehicle. As you note, this may not be realistic, and we may want to be able to model situations where there are different survival functions for each vehicle type. However, this becomes more complicated when both vehicles arrive, but at different times - there will be an interaction between the functions that requires more investigation and is beyond the scope of this study. These scenarios are interesting, and we have added a sentence in the paper explaining this as further work.

\item Reviewer 2 wrote:
\begin{quote}
``Have you given any consideration to using an exact method to solve the optimisation problem solved in Algorithm 1? Is there any alternative in the literature?''
\end{quote}
We did not consider using an exact method to solve this as the were so many non-linear elements to the problem, namely the objective function and the utilisation considerations. An exact method would be beyond the scope of this paper, which focusses on the non-linear objective function formulation, and simulation modelling. We have added using exact methods to the future work discussion.

@@

\item Reviewer 2 wrote:
\begin{quote}
``It seems there are two conflicting methods here, the optmisiation with the EA, which uses some averages, etc. And a more comprehensive DES simulation; however the latter is not used for the optimisation. Did you consider simulation optimisation or simheuristics to address this?''
\end{quote}
We did not consider using the simulation as part of the optimisation process, and is beyong the scope of the paper. However, we think this is a great idea, and have discussed this as future work in the paper, as a tool to give greater consideration to the variability in the system.

\item Reviewer 2 wrote:
\begin{quote}
``P15 Paragraph starting L36. Can you give more detail on the modelling of the call arrival rates? For rare specialties, and given the low usage of ambulances in Indonesia, you might not have enough data to give an accurate representation of actual demand. Especially, for future demand where visibility is increased, how will this change for example among richer/poorer parts of the city? I acknowledge it is impossible to include everything in your experiments, but how do you expect these limitations might affect your results?''
\end{quote}
The call arrival rates are taken directly from provided data, and reported in the Appendix. You make a good point regarding the pick-up locations with particularly low arrival rates - this is not a problem for \textbf{D13} as this is what was observed, but might result in artificially low arrival rates when multiplying up by demand. This is now discussed in the paper.

\item Reviewer 2 wrote:
\begin{quote}
``Could you conduct any validation of your results with real data from eg. Ambulance response times with the baseline allocation?''
\end{quote}
Thorough verification of the models is done by the automated code testing suits provided in the code base, a paragraph on this is now included under Section 5.5. We validated the model based on expert opinion from our collaborators in Jakarta at 118 and 119, and several workshops have been carried out in Indonesia where the model was shared. This is now discussed in Section 6.

### 2024-08-12 tex/main.tex
addresses comments from Vince and Daniel

Exact synchronicity considerations are shown in Figure~\ref{fig:sequential_logic}:
\begin{itemize}
  \item whenever the secondary vehicle reaches the patient before the primary
  has left, vehicle, they remain with the patient until the primary vehicle
  leaves;
  \item if secondary vehicle arrives after the primary vehicle has left, they
  will immediately return to their stations as they are not needed at the scene;
  \item if the expected time for the secondary vehicle to reach the patient
  exceeds the expected time for the primary vehicle to reach the patient then
  the secondary vehicle is not deployed, and so that transit job would be
  abandoned in the secondary simulation (although still seen by a primary
  vehicle in the initial simulation). We assume that secondary vehicles must
  still reach the pickup locations, due to potential difficulties in
  communication en-route. \textcolor{red}{Consider also the situation when a
  secondary vehicle arrives and the patient does not require a hospital, with
  better en-route communication this would mean the primary vehicle turns
  around, however without en-route communication which is assumed here, the
  primary vehicle must still reach the patient pick-up location.}
\end{itemize}

@@

\textcolor{red}{
Verification of a model is ensuring that the conceptual model is implemented
correctly and without error, while validation of a model ensures that the
conceptual model accurately represents the real system to be modelled
\citep{robinson2014simulation}. Verifying the simulation code, alongside
verifying the optimisation and MESLMHPHF objective function code, is done
through automated code testing \citep{percival2014test, palmer2023implementing}.
This includes unit testing to ensure that each sub-component works are requires,
and is bug and error free; end to end testing, ensuring that the components
interact together as intended; and extreme value testing ensuring that changes
in input parameters result in sensible and expected changes in output results
and behaviours. The testing suits are included in the open code repositories
provided.
}

@@

All models are parameterised using demand data that covers all 261
neighbourhoods from 1 January to 31 December 2019, before the COVID-19
pandemic. Data from 2020-2021 was naturally heavily skewed by the pandemic,
with significantly lower demand, and so was not considered as representative
of a typical period for forecasting future needs, hence the decision to use
the available 2019 demand. \textcolor{red}{APPENDIX}~\ref{apx:parameters}
provides further details on the parameterisation of the model, including how
travel time estimates were obtained.
\textcolor{red}{Verification of the models were discussed in
Section~\ref{sec:combined_model}, however parameter validation for this
particular case study is difficult, with all primary data available to us being
used to parametrise the model, still leaving many aspects to be determined by
expert opinion. This expert opinion however was vital in giving a basic
validation of the model, with staff from Ambulans 118 and the 119 Emergency
Ambulance Service in Jakarta approving model results parametrised from their
provided data and expert opinion.}

@@

Table~\ref{tbl:current_optimal_compare} compares the MESLMHPHF objective
function value and simulation derived KPIs between the current and the optimised
allocations, for each demand scenario. It can be seen that, as expected, as
demand increases then utilisation of primary and secondary vehicles increase, as
the ambulance service is busier. Similarly, mean response time increases with
demand, this is due to the vehicles being busier, and therefore less chance that
the most optimally placed vehicle is dispatched. This is confirmed by the
percentage of calls abandoned increasing with demand, showing ambulance
unavailability.
Looking at the expected survival, that is the value of the MESLMHPHF objective
function, we see that using an better allocation does increase the survival,
and that the drop in survival due to demand increases is not as severe when
using an improved allocation in comparison to the current allocation. We can
also see that in particular A1 patients are benefiting from the improved
allocations, as these are the patients most sensitive to small fluctuations in
response time, due to their survival function Equation~\ref{eqn:survival_curve}.
\textcolor{red}{There is a large discrepancy between the expected survival
measures calculated from the MESLMHPHF objective, and the observed survival from
the simulation. This is discussed further in the next section.}
It is worth noting that for the simulation based KPIs (mean response times and
percent abandoned), there is not too much
difference between the currently used allocation and the better allocation,
showing that these KPIs may not be good approximations of survival.

@@

\textcolor{red}{
The discrepancies are exacerbated by the use of cut-off survival function curves
for most of the patients: consider a survival function with cut-off $L$, with
response time below $L$ the patient's chance of survival is 0, and above $L$ the
patient's chance of survival is guaranteed. The optimisation will find a
solution aim to push the expected response time above $L$, and once this is
reached, possibly with an expected response time as low as $L$ itself, it will
not attempt to push the expected
response time above this, as it will have no effect on the expected survival.
However considering variability of mean response times, a considerable
proportion of the response times will lie below the expected value, and so
possibly below $L$, resulting in worse survival probabilities, as observed in
Table~\ref{tbl:current_optimal_compare}. Therefore the distributions chosen to
model travel times can have a large impact on the survival measures, in this
case the naive Triangular distribution.}

@@

\textcolor{red}{In building the models, some assumptions or relaxations were
made to simplify the model, and future work should investigate the effects of
these, for example the assumption that service times are not location dependent
is unrealistic, but relaxed here. Further developments of the model could
consider explicitly modelling service rates using the travel time matrices.
Another limitation identified was the lack of consideration of variability in
travel times in the optimisation model, which cased some discrepancies between
the simulated and expected survivals. One method to overcome this could be to
incorporate variability directly into the MESLMHPHF objective, for example using
percentiles instead of averages, or to optimise using the simulation as an
objective. However in both approaches, a good estimate of travel time
distributions would be needed, with better validation.
}

>>>>

Exact synchronicity considerations are shown in Figure~\ref{fig:sequential_logic}:
\begin{itemize}
  \item whenever the secondary vehicle reaches the patient before the primary
  has left, vehicle, they remain with the patient until the primary vehicle
  leaves;
  \item if secondary vehicle arrives after the primary vehicle has left, they
  will immediately return to their stations as they are not needed at the scene;
  \item if the expected time for the secondary vehicle to reach the patient
  exceeds the expected time for the primary vehicle to reach the patient then
  the secondary vehicle is not deployed, and so that transit job would be
  abandoned in the secondary simulation (although still seen by a primary
  vehicle in the initial simulation). We assume that secondary vehicles must
  still reach the pickup locations, due to potential difficulties in
  communication en-route. \textcolor{red}{Consider a situation when a
  secondary vehicle arrives and the patient does not require a hospital, with
  better en-route communication this would mean the primary vehicle turns
  around, however without en-route communication which is assumed here, the
  primary vehicle must still reach the patient pick-up location.}
\end{itemize}

@@

\textcolor{red}{
Verification of a model is ensuring that the conceptual model is implemented
correctly and without error, while validation of a model ensures that the
conceptual model accurately represents the real system to be modelled
\citep{robinson2014simulation}. Verifying the simulation code, alongside
verifying the optimisation and MESLMHPHF objective function code, is done
through automated code testing \citep{percival2014test, palmer2023implementing}.
This includes unit testing to ensure that each sub-component works are requires,
and is bug and error free; end to end testing, ensuring that the components
interact together as intended; and extreme value testing ensuring that changes
in input parameters result in sensible and expected changes in output results
and behaviours. The testing suites are included in the open code repositories
provided.
}

@@

All models are parameterised using demand data that covers all 261
neighbourhoods from 1 January to 31 December 2019, before the COVID-19
pandemic. Data from 2020-2021 was naturally heavily skewed by the pandemic,
with significantly lower demand, and so was not considered as representative
of a typical period for forecasting future needs, hence the decision to use
the available 2019 demand. \textcolor{red}{APPENDIX}~\ref{apx:parameters}
provides further details on the parameterisation of the model, including how
travel time estimates were obtained.
\textcolor{red}{Verification of the models were discussed in
Section~\ref{sec:combined_model}, we validated based on expert opinion from
our collaborators, Ambulans 118 and the 119 Emergency Ambulance Service in
Jakarta, where several workshops have been carried out in Indonesia where the
model was shared and discussed. This expertise was vital in giving a basic
validation of the model, with collaborators approving model results
parametrised from their provided data and expert opinion.}

@@

Table~\ref{tbl:current_optimal_compare} compares the MESLMHPHF objective
function value and simulation derived KPIs between the current and the optimised
allocations, for each demand scenario. It can be seen that, as expected, as
demand increases then utilisation of primary and secondary vehicles increase, as
the ambulance service is busier. Similarly, mean response time increases with
demand, this is due to the vehicles being busier, and therefore less chance that
the most optimally placed vehicle is dispatched. This is confirmed by the
percentage of calls abandoned increasing with demand, showing ambulance
unavailability.
Looking at the expected survival, that is the value of the MESLMHPHF objective
function, we see that using an better allocation does increase the survival,
and that the drop in survival due to demand increases is not as severe when
using an improved allocation in comparison to the current allocation. We can
also see that in particular A1 patients are benefiting from the improved
allocations, as these are the patients most sensitive to small fluctuations in
response time, due to their survival function Equation~\ref{eqn:survival_curve}.
\textcolor{red}{There is a discrepancy between the expected survival
measures calculated from the MESLMHPHF objective, and the observed survival from
the simulation. This is discussed further in the next section.}
It is worth noting that for the simulation based KPIs (mean response times and
percent abandoned), there is not too much
difference between the currently used allocation and the better allocation,
showing that these KPIs may not be good approximations of survival.

@@

\textcolor{red}{
The discrepancies are exacerbated by the use of cut-off survival function curves
for most of the patients: consider a survival function with cut-off $L$, with
response time below $L$ the patient's chance of survival is 0, and above $L$ the
patient's chance of survival is guaranteed. The optimisation will find a
solution aiming to push the expected response time above $L$, and once this is
reached, possibly with an expected response time as low as $L$ itself, it will
not attempt to push the expected
response time above this, as it will have no effect on the expected survival.
However considering variability of mean response times, a considerable
proportion of the response times will lie below the expected value, and so
possibly below $L$, resulting in worse survival probabilities, as observed in
Table~\ref{tbl:current_optimal_compare}. Therefore the distributions chosen to
model travel times can have a large impact on the survival measures, in this
case the naive Triangular distribution.}

@@

\textcolor{red}{In building the models, some assumptions or relaxations were
made to simplify the model, and future work should investigate the effects of
these, for example the assumption that service times are not location dependent
is unrealistic, but relaxed here. Further developments of the model could
consider explicitly modelling service rates using the travel time matrices.
Another limitation identified was the lack of consideration of variability in
travel times in the optimisation model, which cased some discrepancies between
the simulated and expected survivals. One method to overcome this could be to
incorporate variability directly into the MESLMHPHF objective, for example using
percentiles instead of averages, or to optimise using the simulation as an
objective. However in both approaches, a good estimate of travel time
distributions would be needed, with better validation.
Another direction of future work is to find exact methods of solving the
MESLMHPHF instead of relying on a metaheuristic algorithm.}
